---
title: "ProteomicAnalysis_UnwinECandHIPdata"
author: "DG"
date: "2026-04-17"
output: html_document
toc: true
toc_depth: 2
theme: flatly
---
#The code below demostrates analysis of protein abundance values for AD and control samples. Proteomic data were obtained from a publicly available Alzheimer's disease meta-analysis dataset, specifically the Unwin et al. cohort (entorhinal cortex and hippocampus). 
---
## 1.1 Install Packages 
install.packages("BiocManager")

BiocManager::install("clusterProfiler","limma","org.Hs.eg.db")

install.packages(c("ggplot2", "dplyr", "ggrepel"))

install.packages("VennDiagram")

install.packages("futile.logger")

install.packages("pheatmap")

## 1.2 Load Libraries 
library(clusterProfiler)

library(org.Hs.eg.db)

library(dplyr)

library(ggplot2)

library(ggrepel)  # Optional but recommended for clean labeling

library(tibble)

library(VennDiagram)

library(grid)
library(pheatmap)

## 2. Data Preprocessing
##Load Data

### download data (data17_UnwinHIP_formatted.txt and data18_UnwinEC_formatted.txt) from a github repository (https://github.com/hazhay/Meta-analysis_AD/tree/main/Data)

unwin_ec <- read.delim("data18_UnwinEC_formatted.txt", stringsAsFactors = FALSE, check.names = FALSE)

unwin_hip <- read.delim("data17_UnwinHIP_formatted.txt", stringsAsFactors = FALSE, check.names = FALSE)

## 2.1. Inspect Data
head(unwin_ec)

dim(unwin_ec)

head(unwin_hip)

dim(unwin_hip)

## 2.2. Check NA levels per sample
colSums(is.na(unwin_ec))

colSums(is.na(unwin_hip))

## 2.3. Check distribution across all samples
boxplot(unwin_ec, las=2, main="Sample distributions")

## Sample Distributions EC ##
```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/SampleDistribution-Unwin_EC.png")
```
<img src="`r img_data`" style="width:70%; display:block; margin:auto;" />


boxplot(unwin_hip, las=2, main="Sample distributions")

## Sample Distributions HIP ##
```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data1 <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/SampleDistribution_Unwin-HIP.png")
```
<img src="`r img_data1`" style="width:70%; display:block; margin:auto;" />

## 2.4. Compare control vs AD – biological check
boxplot(unwin_ec[, group1 == "Control"],
        unwin_ec[, group1 == "AD"],
        names=c("Control","AD"))
        
```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data2 <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/Boxplot-Unwin-EC.png")
```
<img src="`r img_data2`" style="width:70%; display:block; margin:auto;" />

boxplot(unwin_ec[, group2 == "Control"],
        unwin_ec[, group2 == "AD"],
        names=c("Control","AD"))

```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data3 <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/Boxplot-Unwin-HIP.png")
```
<img src="`r img_data3`" style="width:70%; display:block; margin:auto;" />

## 2.5. Confirm the raw data structure
readLines("data18_UnwinEC_formatted.txt", n = 2)

readLines("data17_UnwinHIP_formatted.txt", n = 2)

cat(readLines("data18_UnwinEC_formatted.txt", n = 1))

cat(readLines("data17_UnwinHIP_formatted.txt", n = 1))

## 2.6.1. Force a strict delimiter parser
unwin_ec <- read.table(
  "data18_UnwinEC_formatted.txt",
  header = TRUE,
  sep = "",          # auto-detect ANY whitespace 
  stringsAsFactors = FALSE,
  check.names = FALSE
)


unwin_hip <- read.table(
  "data17_UnwinHIP_formatted.txt",
  header = TRUE,
  sep = "",          # auto-detect ANY whitespace 
  stringsAsFactors = FALSE,
  check.names = FALSE
)

## 2.6.2. Ensure proper grouping
group1 <- ifelse(grepl("^C", colnames(unwin_ec)), "Control",
+                  ifelse(grepl("^AD", colnames(unwin_ec)), "AD", NA))

group2 <- ifelse(grepl("^C", colnames(unwin_hip)), "Control",
+                  ifelse(grepl("^AD", colnames(unwin_hip)), "AD", NA))


## 2.6.3. Verify proper grouping
table(group1)

dim(ec)

length (group1)

table(group2)

dim(hip)

length (group 2)

## 2.7. Run Principal Component Analysis (PCA) - QC step

pca <- prcomp(t(ec), scale.=TRUE)

plot(pca$x[,1], pca$x[,2],
     col=group1,
     pch=19,
     xlab="PC1",
     ylab="PC2",
     main="PCA - Entorhinal Cortex")

legend("topright", legend=levels(group1), col=1:2, pch=19)


```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data4 <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/PCA-ECdataset.png")
```
<img src="`r img_data4`" style="width:70%; display:block; margin:auto;" />

pca <- prcomp(t(hip), scale.=TRUE)

plot(pca$x[,1], pca$x[,2],
     col=group2,
     pch=19,
     xlab="PC1",
     ylab="PC2",
     main="PCA - Hippocampus")

legend("topright", legend=levels(group2), col=1:2, pch=19)

```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data5 <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/PCA-HIPdataset.png")
```
<img src="`r img_data5`" style="width:70%; display:block; margin:auto;" />

## 3. Limma Analysis

## 3.1. Design matrix
design <- model.matrix(~group1)

design <- model.matrix(~group2)

## 3.2. Fit model
fit <- lmFit(ec, design)

fit <- eBayes(fit)

fit <- lmFit(hip, design)

fit <- eBayes(fit)

## 3.3. Extract Results
results_ec <- topTable(fit, coef=2, number=Inf)

head(results_ec)

results_hip <- topTable(fit, coef=2, number=Inf)

head(results_hip)

## 3.4. Save Results 
write.csv(data.frame(Gene = rownames(results), results), "limma_results_ec.csv", row.names = FALSE)

write.csv(data.frame(Gene = rownames(results), results), "limma_results_hip.csv", row.names = FALSE)

## 4. Draw plots, graphs and heatmaps

## 4.1. Volcano Plot for EC dataset
results_ec$significant <- results_ec$adj.P.Val < 0.05

ggplot(results_ec, aes(x=logFC, y=-log10(P.Value))) +
  geom_point(aes(color=significant)) +
  scale_color_manual(values=c("grey", "red")) +
  theme_minimal() +
  ggtitle("Volcano Plot - EC")

```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data6 <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/VolcanoPlot-ECdataset.png")
```
<img src="`r img_data6`" style="width:70%; display:block; margin:auto;" />


## 4.2. Volcano Plot for HIP dataset
results_hip$significant <- results_hip$adj.P.Val < 0.05

ggplot(results_hip, aes(x=logFC, y=-log10(P.Value))) +
  geom_point(aes(color=significant)) +
  scale_color_manual(values=c("grey", "blue")) +
  theme_minimal() +
  ggtitle("Volcano Plot - HIP")

```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data7 <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/VolcanoPlot-HIPdataset.png")
```
<img src="`r img_data7`" style="width:70%; display:block; margin:auto;" />

## 4.3. Volcano Plot for EC dataset – TOP Proteins
ggplot(results_ec, aes(x = logFC, y = -log10(P.Value))) +
  geom_point(aes(color = significant)) +
  geom_text_repel(
    data = top_genes,
    aes(label = rownames(top_genes)),
    size = 3
  ) +
  theme_minimal() +
  ggtitle("Volcano Plot - Differentially Abundant Proteins (EC)")
  
```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data8 <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/VolcanoPlot-ECdataset_TopProteins_v2.png")
```
<img src="`r img_data8`" style="width:70%; display:block; margin:auto;" />
  
## 4.4. Volcano Plot for HIP dataset – TOP Proteins
ggplot(results_hip, aes(x = logFC, y = -log10(P.Value))) +
  geom_point(aes(color = significant)) +
  geom_text_repel(
    data = top_genes,
    aes(label = rownames(top_genes)),
    size = 3
  ) +
  theme_minimal() +
  ggtitle("Volcano Plot - Differentially Abundant Proteins (HIP)")

```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data9 <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/VolcanoPlot-HIPdataset_TopProteins_v2.png")
```
<img src="`r img_data9`" style="width:70%; display:block; margin:auto;" />

## 4.5. Volcano Plots (significance thresholds)

#Step 1 — Define thresholds
pval_cutoff <- 0.05

logfc_cutoff <- 1

#Step 2 — Classify proteins
results_ec$category <- "Not Significant"

results_ec$category[results_ec$adj.P.Val < pval_cutoff & results_ec$logFC > logfc_cutoff] <- "Upregulated"
results_ec$category[results_ec$adj.P.Val < pval_cutoff & results_ec$logFC < -logfc_cutoff] <- "Downregulated"

results_hip$category <- "Not Significant"

results_hip$category[results_hip$adj.P.Val < pval_cutoff & results_hip$logFC > logfc_cutoff] <- "Upregulated"
results_hip$category[results_hip$adj.P.Val < pval_cutoff & results_hip$logFC < -logfc_cutoff] <- "Downregulated"

#Step 3 — Select top proteins to label
top_proteins_ec <- results_ec[order(results_ec$adj.P.Val), ]
top_proteins_ec <- top_proteins_ec[1:10, , drop = FALSE]

top_proteins_hip<- results_hip[order(results_hip$adj.P.Val), ]
top_proteins_hip <- top_proteins_hip[1:10, , drop = FALSE]

#Step 4 — Draw Volcano Plots 

# Volcano Plot for EC dataset - TOP Proteins

ggplot(results_ec, aes(x = logFC, y = -log10(P.Value))) +
  
  # Points
  geom_point(aes(color = category), alpha = 0.7) +
  
  # Color scheme
  scale_color_manual(values = c(
    "Upregulated" = "red",
    "Downregulated" = "blue",
    "Not Significant" = "grey"
  )) +
  
  # Threshold lines
  geom_vline(xintercept = c(-logfc_cutoff, logfc_cutoff), 
             linetype = "dashed", color = "black") +
  
  geom_hline(yintercept = -log10(pval_cutoff), 
             linetype = "dashed", color = "black") +
  
  # Labels
  geom_text_repel(
    data = top_proteins_ec,
    aes(label = rownames(top_proteins_ec)),  # or Gene if using CSV
    size = 3,
    max.overlaps = 20
  ) +
  
  # Theme
  theme_minimal(base_size = 14) +
  
  # Labels
  labs(
    title = "Volcano Plot - Differentially Abundant Proteins (EC)",
    x = "Log2 Fold Change",
    y = "-log10(p-value)",
    color = "Category"
  )
  
```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data10 <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/VolcanoPlot-DAP-ECdataset.png")
```
<img src="`r img_data10`" style="width:70%; display:block; margin:auto;" />
  
# Volcano Plot for HIP dataset – TOP Proteins 

library(ggplot2)
library(ggrepel)

ggplot(results_hip, aes(x = logFC, y = -log10(P.Value))) +
  
  # Points
  geom_point(aes(color = category), alpha = 0.7) +
  
  # Color scheme
  scale_color_manual(values = c(
    "Upregulated" = "red",
    "Downregulated" = "blue",
    "Not Significant" = "grey"
  )) +
  
  # Threshold lines
  geom_vline(xintercept = c(-logfc_cutoff, logfc_cutoff), 
             linetype = "dashed", color = "black") +
  
  geom_hline(yintercept = -log10(pval_cutoff), 
             linetype = "dashed", color = "black") +
  
  # Labels
  geom_text_repel(
    data = top_proteins_hip,
    aes(label = rownames(top_proteins_hip)),  # or Gene if using CSV
    size = 3,
    max.overlaps = 20
  ) +
  
  # Theme
  theme_minimal(base_size = 14) +
  
  # Labels
  labs(
    title = "Volcano Plot - Differentially Abundant Proteins (HIP)",
    x = "Log2 Fold Change",
    y = "-log10(p-value)",
    color = "Category"
  )

```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data11 <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/VolcanoPlot-DAP-HIPdataset.png")
```
<img src="`r img_data11`" style="width:70%; display:block; margin:auto;" />

## 4.6.1. Compare EC vs HIP datasets

plot(results_ec[common,"logFC"],
     results_hip[common,"logFC"])
abline(0,1)

# [1] 0.6939343

```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data12 <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/ECvsHIP-comparison.png")
```
<img src="`r img_data12`" style="width:70%; display:block; margin:auto;" />

## 4.6.2. Define shared significant proteins
sig_ec <- rownames(results_ec[results_ec$adj.P.Val < 0.05, ])
sig_hip <- rownames(results_hip[results_hip$adj.P.Val < 0.05, ])

shared <- intersect(sig_ec, sig_hip)

length(shared)

## 4.6.3. Build plotting dataframe
common <- intersect(rownames(results_ec), rownames(results_hip))

plot_df <- data.frame(
  protein = common,
  logFC_EC = results_ec[common, "logFC"],
  logFC_HIP = results_hip[common, "logFC"]
)

plot_df$category <- ifelse(plot_df$protein %in% shared,
                           "Shared Significant",
                           "Other")

## 4.6.4. Plot comparison

ggplot(plot_df, aes(x = logFC_EC, y = logFC_HIP)) +
  geom_point(aes(color = category), alpha = 0.6) +
  
  scale_color_manual(values = c(
    "Shared Significant" = "red",
    "Other" = "grey"
  )) +
  
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  
  theme_minimal(base_size = 14) +
  
  labs(
    title = "EC vs HIP Protein Fold-Change Correlation",
    x = "EC log2 Fold Change",
    y = "HIP log2 Fold Change",
    color = "Category"
  )

## 4.6.5. Plot comparison with top shared proteins highlighted

top_shared <- plot_df[plot_df$protein %in% shared, ]
top_shared <- top_shared[order(-abs(top_shared$logFC_EC)), ][1:10, ]

ggplot(plot_df, aes(x = logFC_EC, y = logFC_HIP)) +
  geom_point(aes(color = category), alpha = 0.6) +
  
  geom_text_repel(
    data = top_shared,
    aes(label = protein),
    size = 3
  ) +
  
  scale_color_manual(values = c(
    "Shared Significant" = "red",
    "Other" = "grey"
  )) +
  
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  
  theme_minimal(base_size = 14) +
  
  labs(
    title = "Shared Differential Proteins (EC vs HIP)",
    x = "EC log2 Fold Change",
    y = "HIP log2 Fold Change"
  )
  
```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data13 <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/SharedDAP-ECvsHIP.png")
```
<img src="`r img_data13`" style="width:70%; display:block; margin:auto;" />
  
## 4.6.6. Quantify Overlap
length(shared)

#[1] 32

#proportion of shared signal
length(shared) / length(common)

#[1] 0.01369863

## 5. Pathway Enrichment Analysis

## 5.1. Extract Significant Proteins
sig_p <- 0.05
sig_lfc <- 0.25

ec_sig <- results_ec %>%
  as.data.frame() %>%
  rownames_to_column("uniprot") %>%
  filter(adj.P.Val < sig_p, abs(logFC) > sig_lfc)

hip_sig <- results_hip %>%
  as.data.frame() %>%
  rownames_to_column("uniprot") %>%
  filter(adj.P.Val < sig_p, abs(logFC) > sig_lfc)

## 5.2. Map UniProt → Entrez IDs

ec_genes <- bitr(
  ec_sig$uniprot,
  fromType = "UNIPROT",
  toType   = "ENTREZID",
  OrgDb    = org.Hs.eg.db
)

hip_genes <- bitr(
  hip_sig$uniprot,
  fromType = "UNIPROT",
  toType   = "ENTREZID",
  OrgDb    = org.Hs.eg.db
)

## 5.3. Check mapping - QC step

nrow(ec_genes)

#[1] 296

nrow(hip_genes)
#[1] 53

# These numbers suggest that both groups will be relatively stable for GO analysis, but KEGG is expected to be very sparse for the hip dataset. Therefore, continue with GO analysis.

##Additional QC to check protein number before mapping

nrow(ec_sig)
#[1] 301

nrow(hip_sig)
#[1] 52

nrow(ec_genes)
#[1] 296

nrow(hip_genes)
#[1] 53

##This shows that there is good mapping coverage. It's just that the HIP dataset has a much smaller DE signal than EC

## 5.4. Compare clusters
gene_list <- list(
  EC  = ec_genes$ENTREZID,
  HIP = hip_genes$ENTREZID
)

cc_go <- compareCluster(
  geneCluster = gene_list,
  fun = "enrichGO",
  OrgDb = org.Hs.eg.db,
  ont = "BP",
  keyType = "ENTREZID"
)

## 5.5. Create Dotplot
dotplot(cc_go, showCategory = 15, title = "GO Biological Process: EC vs HIP")

```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data14 <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/GObiologicalProcess_ECvsHIP_v1.png")
```
<img src="`r img_data14`" style="width:70%; display:block; margin:auto;" />
  
## 5.6. Identify shared pathways between EC and HIP and plot them

#Step 1 — Get the 9 shared pathways
ec_df  <- as.data.frame(ec_go)
hip_df <- as.data.frame(hip_go)

shared_ids <- intersect(ec_df$ID, hip_df$ID)

shared_df <- ec_df[ec_df$ID %in% shared_ids,
                   c("ID", "Description", "p.adjust")]

#Step 2 — Make sure it’s clean 
shared_df <- na.omit(shared_df)
shared_df <- shared_df[order(shared_df$p.adjust), ]

#Step 3 — Convert to grid table
tbl <- gridExtra::tableGrob(shared_df)

#Step 4 — Save as PNG 
png("shared_pathways.png",
    width = 2200,
    height = 1200,
    res = 300)

grid::grid.draw(tbl)

dev.off()
```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data15 <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/shared_pathways.png")
```
<img src="`r img_data15`" style="width:70%; display:block; margin:auto;" />

## 6. Visualisations (EC-HIP comparison)
##6.1. Heatmap 

#STEP 1 — Build combined pathway table by merging EC + HIP enrichment results
ec_df  <- as.data.frame(ec_go)
hip_df <- as.data.frame(hip_go)
##Keep only shared pathways 
shared_ids <- intersect(ec_df$ID, hip_df$ID)

ec_shared <- ec_df[ec_df$ID %in% shared_ids, ]
hip_shared <- hip_df[hip_df$ID %in% shared_ids, ]

#STEP 2 — Build heatmap matrix by using −log10(p.adjust) 
ec_shared$score  <- -log10(ec_shared$p.adjust)
hip_shared$score <- -log10(hip_shared$p.adjust)
##Combine into one table
library(dplyr)

heat_df <- full_join(
  ec_shared[, c("ID", "Description", "score")],
  hip_shared[, c("ID", "score")],
  by = "ID",
  suffix = c("_EC", "_HIP")
)
#Replace missing values with 0 
heat_df[is.na(heat_df)] <- 0

#Use description as rownames
rownames(heat_df) <- heat_df$Description
heat_matrix <- as.matrix(heat_df[, c("score_EC", "score_HIP")])

#STEP 3 — Create heatmap
pheatmap(
  heat_matrix,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  color = colorRampPalette(c("white", "red"))(100),
  main = "Shared Pathway Enrichment (EC vs HIP)"
)

```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data16 <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/PathwayEnrichmentHeatmap_ECvsHIP.png")
```
<img src="`r img_data16`" style="width:70%; display:block; margin:auto;" />

## 6.2. Venn Diagram

#Step 1 — define the two sets
ec_set  <- ec_genes$ENTREZID
hip_set <- hip_genes$ENTREZID

#Step 2 — create Venn plot
venn.plot <- venn.diagram(
  x = list(
    EC = ec_set,
    HIP = hip_set
  ),
  filename = NULL,
  fill = c("steelblue", "tomato"),
  alpha = 0.5,
  cex = 1.5,
  cat.cex = 1.3,
  cat.pos = 0
)

#Step 3 — display venn plot
grid.newpage()
grid.draw(venn.plot)

```{r, echo=FALSE}
# This converts your local image into a Base64 string and displays it
# No external path will be needed after the HTML is generated
img_data17 <- knitr::image_uri("/Users/mirandapestrofa/Downloads/MultimodalAnalysisAD/ProteomicDatasetAnalysis/VennDiagram-ECvsHIP.png")
```
<img src="`r img_data17`" style="width:70%; display:block; margin:auto;" />


---
